---
title: "DIF with several person factors and repeated measures"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DIF with several person factors and repeated measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Design the analysis around the person

Differential item functioning is a violation of the Rasch model's requirement
of invariant comparison: item locations should not depend on which persons
respond (Rasch 1961). When several person factors are relevant, they should
enter one model. Repeated observations require a further distinction: group is
a between-person factor, whereas occasion varies within person.

For a factor $G$, class interval $C$, and standardised residual $z$, the
single-factor model is

$$
z=\mu+G+C+G\mathbin{:}C+\varepsilon.
$$

The $G$ term tests uniform DIF. The $G\mathbin{:}C$ term tests non-uniform DIF.
With several factors, `dif_anova` fits their terms jointly and uses Type II
sums of squares.

The following dataset has two observations per person. Item I03 has a group
shift, item I06 has an occasion shift, and item I05 shifts for group B at the
second occasion only, a group-by-occasion interaction.

```{r data}
set.seed(21)
N <- 320
difficulty <- seq(-1.5, 1.5, length.out = 8)
theta <- rnorm(N)
group <- rep(c("A", "B"), each = N / 2)

make_wave <- function(occasion_shift, interaction_shift) {
  shift <- matrix(0, N, 8)
  shift[group == "B", 3] <- 1.2
  shift[, 6] <- occasion_shift
  shift[group == "B", 5] <- interaction_shift
  matrix(rbinom(N * 8, 1,
                plogis(outer(theta, difficulty, "-") - shift)), N, 8)
}

X <- rbind(make_wave(0, 0), make_wave(1.0, 2.0))
colnames(X) <- sprintf("I%02d", 1:8)
dat <- data.frame(
  pid = rep(sprintf("P%03d", seq_len(N)), 2),
  X,
  group = rep(group, 2),
  occasion = rep(c("T1", "T2"), each = N)
)

# the same three persons at each occasion: the identifier repeats down the
# rows, which is what makes the design repeated measures
dat[c(1:3, N + 1:3), c("pid", "I01", "I02", "I03", "group", "occasion")]
```

## Fit once and test both factors

The repeated person identifier is carried into the fit. `dif_anova` detects
occasion as within-person; it can also be declared explicitly. Persons, not
stacked rows, are the units of analysis. Uniform between-person terms use
Type II tests with HC3 covariance. Class-interval interactions retain the
residual-ANOVA reference. Within-person tests use person-level contrasts, with
a Greenhouse--Geisser correction when a factor has more than two levels.
This mixed-design analysis extends the single-factor residual analysis of
variance described by Andrich and Marais (2019). Its F references are
large-sample approximations.

With incomplete panels, between-person tests fit the occasion and person
factors jointly, giving each person total weight one. They use person-cluster
CR3 covariance for both uniform and non-uniform terms, including uncertainty
in the occasion adjustment. This avoids confusing unequal occasion coverage
with differences in the composition of the person groups. Within-person
contrasts still require complete panels.

```{r analysis}
fit <- rasch(dat, id = "pid", factors = c("group", "occasion"),
             items = sprintf("I%02d", 1:8))
da <- dif_anova(fit, within = "occasion", effects = "factorial", sizes = TRUE)
da$summary
```

The multiplicity adjustment covers the complete family of item-by-DIF-term
tests. Uniform DIF is a factor effect that is stable over the trait; a
factor-by-class-interval effect is non-uniform DIF. `effects = "factorial"`
adds the person-factor interactions. A significant higher-order term
supersedes its component group terms within the same item: item I05's group
effect is significant on its own, but the `superseded` flag records that the
interaction absorbs it, and the follow-ups report the interaction rather than
its components. Read adjusted probabilities with effect sizes before changing
an item.

`dif_bootstrap()` repeats this design under the fitted invariant model. Ordinary
and explanatory Rasch models and Multiple Ratings condition on each person's
score. Extended Frames conditions within each item set, where the frame unit is
common. Comparative Judgement instead draws outcomes from its fitted model
while retaining judges and the comparison design. Every replicate repeats the
calibration and the full DIF family. The result is a sensitivity analysis; the
adjusted residual ANOVA remains primary. Its familywise probabilities describe
the fitted global invariant null, not otherwise invariant items when another
item has DIF.

```{r bootstrap-sensitivity, eval = FALSE}
db <- dif_bootstrap(fit, da, B = 999, workers = 4, seed = 2026)
db$summary[, c("item", "term", "p_uniform_boot_adj",
               "p_nonuniform_boot_adj")]
```

In a four-occasion validation study, familywise error under the balanced
Rasch null was 6.9% for the primary analysis and 4.9% for the bootstrap over
350 datasets. For uniform occasion DIF the bootstrap was
slightly less powerful than the primary analysis. Adjusted non-uniform power
was weak for both procedures in this seven-item, 180-person design. The
bootstrap therefore remains a sensitivity analysis rather than the default
test.

The earlier panel-loss results used the superseded marginal adjustment.
For the current joint adjustment, a known-residual study with correlated
person factors and unequal panels gave uniform and non-uniform null rejection
rates of 5.7% and 5.3% over 3,000 replicates (Monte Carlo SEs 0.42 and 0.41
percentage points). A separate end-to-end Rasch check gave 3.0% over 200
datasets for a null group effect on an item carrying DIF for another factor.
These checks assess the primary tests, not the full bootstrap family's error
rate under every incomplete-panel design.

## Quantify the departure

ANOVA identifies evidence against invariance; it does not state the size of
the departure in logits. `dif_size` resolves an item by a between-person
factor. `dif_contrasts` provides planned contrasts and uses person-level
differencing for within-person questions.

```{r magnitude}
dif_size(fit, "I03", by = "group")
dc <- dif_contrasts(fit, items = c("I03", "I06"), within = "occasion")
dc$table
da$posthoc
```

`dif_posthoc()` is the general follow-up for a significant term. A main
effect with more than two levels is reported as pairwise marginal differences
over the other fitted factors. An interaction is reported as a
difference-in-differences, or its higher-order counterpart. These comparisons
use the joint covariance of the resolved item locations and Holm adjustment
over the stated family. The result is on the logit scale and respects the
factor structure used in the DIF analysis. Here item I03 is reported as a
pairwise group difference, item I06 as a person-level occasion contrast, and
item I05 as the difference-in-differences of its interaction; the superseded
I05 group term receives no follow-up of its own.

For repeated-person contrasts, significance comes from person-level residual
contrast scores with the same design-cell weights as the resolved estimate.
Other fitted factors are averaged equally over their complete cells, including
when their sample sizes differ, and the independent between-person cells use a
Welch--Satterthwaite reference. The resolved logit difference remains the
magnitude. Its standard error uses the person-clustered calibration covariance,
whereas `dif_contrasts()` tests a nominated within-person comparison directly
from person-level contrasts. The fitted person identifier is used unless `id`
is supplied explicitly.

For a many-facet fit, follow-ups may name the underlying item; its virtual
facet cells are pooled with common weights so facet severity cancels from the
group contrast. Extended-frame fits support the residual ANOVA for factors
outside the frame definition, but not an ordinary resolved-item magnitude:
that refit would discard the fitted frame units.

A statistical flag should be considered with the logit magnitude, targeting,
item content, and the intended use of the scale. Resolving an item changes the
measurement model and should follow a substantive account of why the item is
not invariant.

## References

Andrich, D., and Marais, I. (2019). *A Course in Rasch Measurement Theory:
Measuring in the Educational, Social and Health Sciences*. Springer.

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). Berkeley: University of
California Press.
