---
title: "Explanatory Rasch models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explanatory Rasch models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rasch)
```

The linear logistic test model (LLTM; Fischer, 1973) explains dichotomous item
locations using observed item characteristics. The linear partial credit model
(LPCM; Fischer and Ponocny, 1994) extends the same formulation to polytomous
thresholds. Both retain the common discrimination and conditional Rasch
structure.

For category \(k\) of item \(i\),

\[
\log\frac{P(X_{ni}=k)}{P(X_{ni}=k-1)}
=\theta_n-\delta_{ik}, \qquad
\delta_{ik}=\mathbf z_{ik}^{\mathsf T}\gamma .
\]

The vector \(\mathbf z_{ik}\) contains the nominated item or threshold
characteristics. An interaction permits the effect of one characteristic to
differ according to another. The scale origin is fixed at mean item location
zero, as in `rasch()`; the formula intercept is therefore absorbed by the
origin constraint.

Design columns are centred and rescaled internally for numerical stability. Coefficients
and standard errors are reported in the supplied predictor units, for both
Rasch and comparative judgement models.

## Item predictors

Item-level metadata contain one row per item. This example explains eight
dichotomous item locations by the operation required and the response format.

```{r lltm}
set.seed(1)
item_design <- data.frame(
  item = paste0("I", 1:8),
  operation = rep(c("recall", "inference"), each = 4),
  format = rep(c("selected", "constructed"), 4)
)

difficulty <- 0.8 * (item_design$operation == "inference") +
  0.4 * (item_design$format == "constructed")
group <- factor(rep(c("A", "B"), each = 250))
theta <- rnorm(500) + ifelse(group == "B", 0.3, 0)
X <- sapply(difficulty, function(delta)
  rbinom(length(theta), 1, plogis(theta - delta)))
colnames(X) <- item_design$item

fit <- rasch_explanatory(
  X,
  predictors = item_design,
  formula = ~ operation + format,
  level = "item",
  factors = data.frame(group = group)
)
fit$est$coefficients
```

If the same person contributes more than one response row, the coefficient
covariance is clustered by person. Supported repeated-person fits use a
linearised delete-one-person covariance correction. Every supported fit takes
a \(t\) reference with degrees of freedom equal to the number of independent
person units contributing conditional information minus one, whether or not
identifiers repeat: one response row per person still leaves a finite count of
sampling units, and reading that count as a limiting normal rejects a true
null more often than the printed probability states. Coefficient inference is
withheld when the calibration lacks enough independent information. Holm
adjustment covers the coefficient family.
With few persons and unequal numbers of response rows, these approximate tests
can still be mildly liberal. The correction does not guarantee nominal
small-sample coverage.

Predictors may be continuous, categorical or ordinal. Numeric vectors are
continuous and unordered factors are categorical. An ordered factor is scored
with successive contrasts: each coefficient is the change between two
adjacent levels in the declared order. Character vectors are converted to
unordered factors; the first factor level is the reference category. For
example:

```{r predictor-types, eval=FALSE}
item_design$demand <- as.numeric(item_design$demand)
item_design$format <- factor(item_design$format)
item_design$complexity <- ordered(
  item_design$complexity,
  levels = c("low", "moderate", "high")
)
```

This coding uses the order without assuming equal spacing and does not fit
polynomial contrasts. Use an unordered factor for comparisons with a single
reference category, or a numeric predictor when the spacing has a substantive
scale. Selected interactions can be added, for example
`~ demand + format + complexity + format:complexity`.
Formula offsets (`offset()`) are not supported in explanatory Rasch or CJ models.

## Threshold predictors

For an LPCM, item metadata may still be supplied once per item. The reserved
factor `threshold` identifies the within-item threshold and can be used in the
formula:

```{r lpcm-form, eval=FALSE}
lpcm <- rasch_explanatory(
  responses,
  predictors = item_design,
  formula = ~ operation + format + threshold + format:threshold,
  level = "item"
)
```

Use `level = "threshold"` when a predictor varies within an item. The metadata
must then contain `item`, `threshold`, and one row for every observed item
threshold. `threshold_number` is also available as an integer predictor.

## Comparing the restriction with a free calibration

`rasch_explanatory()` retains a free partial credit calibration of the same
responses. `explanatory_test()` compares it with the active explanatory model.
Pairwise conditional estimation produces a composite likelihood, so the
reported inferential result is the first-order Kent-adjusted chi-square rather
than an ordinary likelihood-ratio test. The same table reports calibration
\(R^2\), the proportion of variation in the free threshold calibration
reproduced by the explanatory thresholds. Comparative judgement uses the
corresponding free object calibration. The Kent comparison is asymptotic and
does not use the finite-person-cluster correction applied to individual
coefficients.

```{r comparison}
explanatory_test(fit)
```

A non-significant result does not establish that the explanatory structure is
true. It indicates that the free calibration has not detected a departure at
the available precision. Coefficients should be interpreted together with
their standard errors, item fit, targeting, and the study design.

## Fixed departures

`rasch_explanatory()` begins with the exact explanatory restriction. Its
error formulation uses nominated fixed departures rather than a random item
effect, so it does not introduce an item population distribution. An
explanatory model can be too restrictive for a small number of items.
`explanatory_diagnostics()` adds each available departure separately and
adjusts the complete family of probabilities by Holm's method.

```{r diagnostics}
departures <- explanatory_diagnostics(fit)
head(departures)
```

An item-location departure moves all thresholds of an item together. A
threshold-structure departure changes their relative locations and is
available only for polytomous items. A departure should be accepted on
substantive grounds, not selected solely because it has the smallest
probability.

```{r relax, eval=FALSE}
fit <- relax_explanatory(fit, item = "I4", component = "location")
```

The refit remains a fixed-effects Rasch model: no item distribution or random
effect is introduced. Item locations, person measures, residuals, fit
statistics and subsequent analyses are all recomputed. Item deletion, DIF
splitting and superitem construction retain the explanatory structure and add
the required fixed departures for changed items. Response-dependence
resolution does the same for its resolved item copies. With keyed
multiple-choice data, unchanged and split items retain their raw responses for
distractor analysis.

When person factors are supplied to `rasch_explanatory()`, DIF is tested in
the usual way. The optional bootstrap conditions on each person's score and
refits the active explanatory model, including accepted fixed departures. Its
familywise probabilities refer to the fitted global invariant null.

```{r dif-bootstrap, eval=FALSE}
dif <- dif_anova(fit)
dif_bootstrap(fit, dif, B = 999, seed = 2026)$summary
```

The same workflow is available under **Explanatory** in the Shiny application.
Predictor type is selected explicitly in the application; ordinal predictors
also require their level order. The project file records the predictor
metadata, formula and accepted departures, and the displayed R code reproduces
the analysis.

## Comparative judgement

`btl_explanatory()` constrains object rather than item locations. Predictor
metadata contain one row per object and an `object` column. The formulation is

\[
\log\frac{P(a\succ b)}{P(b\succ a)}
=\beta_a-\beta_b, \qquad
\beta_i=\mathbf z_i^{\mathsf T}\gamma .
\]

```{r cj, eval=FALSE}
cj <- btl_explanatory(
  comparisons,
  predictors = object_design,
  formula = ~ domain + format + domain:format,
  object_a = "object_a",
  object_b = "object_b",
  winner = "winner",
  judge = "judge"
)
explanatory_test(cj)
explanatory_diagnostics(cj)
```

An ordered `response` may be supplied instead of `winner`. Its symmetric
response thresholds are estimated jointly with the explanatory object effects.
Judge-clustered covariance and its existing availability rules continue to
apply. `relax_btl_explanatory()` adds a nominated fixed object-location
departure and repeats the complete fit.

## Validation

Regression checks compare the LLTM and LPCM calibrations with `eRm`. In the
simulation study, coefficient bias was at most
0.005 logits, empirical standard deviations divided by mean reported standard
errors ranged from 0.99 to 1.03, and 95% coverage ranged from 0.938 to 0.955
across LLTM, LPCM, dichotomous comparative judgement, ordered comparative
judgement and judge-clustered comparative judgement conditions. Kent-adjusted
null rejection ranged from 4.2% to 6.0% (1,000 replicates per condition).
Holm familywise error for fixed-departure diagnostics was 4.3% to 4.7%, with
98.3% to 100% power for a planted 0.8-logit departure (300 replicates per
condition). A further 1,000-replicate study covered mixed maximum scores and
four-category LPCMs at 300 to 2,000 persons. SE ratios were 0.993 to 1.026,
coverage was 0.942 to 0.954, Kent-adjusted null rejection was 4.3% to 5.8%,
and mean calibration \(R^2\) was 0.872 to 0.994. The unscaled probability
rejected 98.6% to 100% and is therefore returned only as `p_naive`. No fit in
these four conditions was refused or failed to converge. Scripts and full
result tables are under `tools/simval/` in the source repository.

## References

Fischer, G. H. (1973). The linear logistic test model as an instrument in
educational research. *Acta Psychologica*, 37(6), 359--374.

Fischer, G. H., and Ponocny, I. (1994). An extension of the partial credit
model with an application to the measurement of change. *Psychometrika*,
59(2), 177--192.

Mair, P., and Hatzinger, R. (2007). Extended Rasch modeling: The eRm package
for the application of IRT models in R. *Journal of Statistical Software*,
20(9), 1--20. doi:10.18637/jss.v020.i09.
