---
title: "The Extended Frame of Reference model"
author: "Josh McGrane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Extended Frame of Reference model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r library}
library(rasch)
```

## Frames and units

Rasch's criterion of invariant comparison is stated within a frame of
reference (Rasch 1961). The extended frame of reference model allows the unit
to differ over linked combinations of item sets and person groups (Humphry
2005; Humphry and Andrich 2008). For item \(i\) in set \(s\) and person
\(n\) in group \(g\),

\[
P(X_{ni}=x)=
\frac{\exp\{\rho_{sg}[x\theta_n-\sum_{k=1}^{x}\delta_{ik}]\}}
{\sum_{y=0}^{m_i}\exp\{\rho_{sg}[y\theta_n-
\sum_{k=1}^{y}\delta_{ik}]\}},\qquad
\rho_{sg}=\alpha_s\phi_g .
\]

Following Humphry and Andrich (2008, eq. 15), the multiplicative parameter
of a frame is the ratio of the common reference unit to the frame's own unit:
\(\alpha_s\) carries that ratio for item set \(s\), and \(\phi_g\) for
person group \(g\). Their geometric means are fixed at one; no observed set
or group is the reference. A set with \(\alpha_s = 1.3\) has a smaller
natural unit than the geometric-mean set unit, so its response curves are
steeper on the common scale. Two observed sets can be compared directly by
the ratio \(\alpha_s/\alpha_t\). The partial credit model holds within each
frame in its natural unit; dichotomous items are the one-threshold case.

Use this model when variation in the unit is part of the measurement account.
Poor fit to an equal-unit model is not by itself evidence for a
frame-dependent unit.

## Fit the model

```{r simulate}
d <- simulate_efrm(
  n_per_group = 150,
  items_per_set = 8,
  set_unit_ratio = 1.30,
  group_unit_ratio = 1.10,
  seed = 25
)
truth <- attr(d, "truth")
d$site <- rep(c("A", "B"), length.out = nrow(d))
d
```

The simulated data carry the units the fit below has to recover: a set-unit
ratio of 1.30 and a group-unit ratio of 1.10.

```{r fit}
fit <- rasch_efrm(
  d,
  item_sets = truth$item_sets,
  groups = "group",
  id = "id",
  factors = "site",
  boot_reps = 50,
  workers = 1,
  seed = 25
)
fit
```

The principal model-specific tables are:

```{r tables}
fit$phi_table       # person-group units
fit$alpha_table     # item-set unit ratios (reference unit / set unit)
fit$set_table       # linked set locations
fit$frames          # complete frame units
fit$linking         # set-linking design
```

The frame-defining group is part of the model and is not retested as DIF.
Other person factors can be examined with `dif_anova()`. Its bootstrap
conditions on each person's subtotal within each observed item set, then
repeats the Extended Frames fit and the complete DIF analysis. Its familywise
probabilities refer to the fitted global invariant null.

```{r dif-bootstrap, eval = FALSE}
dif <- dif_anova(fit)
dif_bootstrap(fit, dif, B = 999, seed = 2026)$summary
```

Person-group units are estimated from common threshold patterns across
groups. Set units are estimated from persons observed in more than one set.
For a linked pair, the natural coordinates satisfy
\(u_b=r u_a+c\). The package estimates \(r\), \(c\), and masses for the
nuisance person distribution on a finite grid by semiparametric likelihood:

\[
\prod_n \int P(X_{na}\mid u)P(X_{nb}\mid ru+c)\,dF_{g(n)}(u).
\]

The masses are estimated jointly with the link, separately for each observed
person group, so neither a normal shape nor a common distribution across
groups is prescribed. The link holds the conditional item thresholds and
person-group units fixed while estimating only \(r\), \(c\), and the nuisance
masses. The common-scale parameters are then

\[
\delta_{ik}=\widetilde\delta_{ik}/\alpha_s+\mu_s,
\qquad \rho_{sg}=\alpha_s\phi_g.
\]

The item thresholds and person-group units remain
pairwise conditional estimates; the mixing distribution is used only for
the relative set unit, which is not identified by that conditional stage.
The discrete nonparametric margin follows Follmann (1988); its use for linked
item-set units is an extension implemented here.

The group-by-set frame graph and the set-linking graph must each connect to a
common scale. A linking response pattern must span a score range of at least
four within a set. Overlapping item sets are not permitted.

Several group factors may define crossed frames:

```{r crossed, eval = FALSE}
fit_crossed <- rasch_efrm(
  data,
  item_sets = item_sets,
  groups = c("language", "cohort"),
  id = "id"
)
fit_crossed$phi_factorial
fit_crossed$phi_factorial_tests
```

The factorial tables give the GLS decomposition of the estimated log group
units. Use `p_adj`, which applies Holm's correction across the factorial
terms, for decisions. The tables do not replace the frame-level estimates.

## Uncertainty and equal-unit comparisons

The default hybrid covariance combines the pairwise Godambe covariance with a
person bootstrap for set linking. Within-frame thresholds and group units are
jointly redrawn, and the link is rebuilt from each draw. The resulting
covariance is retained for the common-scale thresholds and complete frame
units.
Set `se_method = "bootstrap"` to refit the complete model to each person
resample.

The linking calculations use a compiled numerical kernel. Bootstrap replicates
run on four workers by default, or fewer where the system limit is lower. A
fixed `seed` gives the same samples and results for every worker count without
altering the caller's random-number stream. In the Shiny application the fit
runs in a separate process. The analyst can select up to four workers, follow
completed batches, or cancel the process without replacing the current
analysis. R scripts can supply `progress` and `cancel` callbacks for the same
controls.

The fit records the requested, usable and failed bootstrap counts. The requested
number must exceed the free directions in the largest covariance block used by
the fit. The stored bootstrap row contains several constrained blocks and is not
treated as one covariance. Once estimation begins, the linking covariance needs
at least 30 usable replicates and a majority of those requested; otherwise the
fit stops. A numerical failure or non-convergence on any supported link
invalidates the whole replicate. Pairs without enough informative common
persons can be omitted only if the remaining links connect all sets.
Full-bootstrap refits with unidentified group units are also discarded.
The conditional calibration checks the exact likelihood curvature as well as
the score. Stationary saddles and flat solutions are refused; the same check
applies to bootstrap refits.
Saved frame fits without a current likelihood-check record must be refitted
before reopening in the app. So must an extended frame fit whose stored score
curves predate the shared design enumeration, since their designs and labels
are no longer the ones `test_information()` and the curve plots would give.
Neither refusal alters the source file.
The alpha--phi cross-covariance has the same usable-draw rule and is
withheld if it cannot be estimated. A full bootstrap that falls below its
minimum returns hybrid standard errors instead, retaining its attempted, usable
and failed counts in `full_boot_reps_*`.

```{r comparison}
fit$efrm_vs_rasch$unit_omnibus
fit$efrm_vs_rasch$unit_tests
```

The omnibus Wald tests assess the set- and group-unit families. Their
probabilities are adjusted together by Holm's method. Individual unit
contrasts form a separate Holm-adjusted follow-up family. The raw
composite-likelihood difference is descriptive and compares the group-unit
stage only. Set units are identified in the linking stage and are assessed by
their Wald test.
`compare_fits()` withholds generic likelihood differences involving EFRM:
its calibration likelihood uses within-set pairs, whereas ordinary PCM also
uses cross-set pairs. The matched group-unit comparison is retained in
`fit$efrm_vs_rasch`.
Probabilities require at least 50 persons or effective persons in every group
and a path between each pair of sets whose links each have at least 50
informative common persons. A person
at the same extreme tail in both sets remains in the likelihood but does not
count as link support. Sparse designs retain the unit estimates without an
inferential probability.

Two-set simulations under normal, bimodal and deliberately different group
distributions gave set-unit bias within 0.004 log-units, empirical-to-reported
SE ratios from 0.97 to 1.05, and null rejection from 4.0 to 5.0 per cent for
the hybrid method. The complete person bootstrap was mildly conservative in
the corresponding null design. Sets with only three dichotomous items were
not identified; four-item sets were recovered without bias in adequately
supported samples.

A separate null study checked the complete decision families. With 200
persons per group and six items per set, 489 of 500 fits were analysed. The
Holm familywise rejection rates were 3.5 per cent for the omnibus tests and
1.6 per cent for the individual unit contrasts. A boundary design with 100
persons per group and four items per set refused 493 of 1,000 fits and did not
converge in another 24. Among the 483 analysed fits, the corresponding rates
were 5.2 and 2.5 per cent. A design should not be used merely because it is
formally identifiable.

## Item invariance across frames

The fitted EFRM gives an item one location across its frames, scaled by the
frame unit. `frame_invariance()` examines that restriction by calibrating
each frame separately. Each compared set and frame needs at least 50 distinct
persons contributing an informative item pair: not both zero or both at their
item maxima after any item removal or category recoding in that frame's
separate calibration. Items with weak separate-frame standard errors are excluded
from the comparison.

For frame \(f\), the common-scale item location is
\(\hat\delta_{if}^{*}=\hat\delta_{if}/\hat\rho_f\). The separate
calibrations have independent origins, so differences are centred over the
common thresholds before testing. Let \(w_i=m_i/\sum_jm_j\), where \(m_i\)
is the number of thresholds for item \(i\). The default covariance conditions
on the fitted frame units:

\[
C\{V_1/\hat\rho_1^2+V_2/\hat\rho_2^2\}C^{\mathsf T},
\qquad C=I-\mathbf 1\mathbf w^{\mathsf T} .
\]

```{r invariance, eval = FALSE}
inv <- frame_invariance(fit, se_method = "conditional")
inv$summary
inv$locations
inv$discrimination
```

A whole-person bootstrap within person group retains each sampled row's
item-set response pattern and includes uncertainty in the fitted units. A
replicate is used only if it retains the same item-comparison family, and
therefore the same centred frame origin, as the observed analysis:

```{r invariance-bootstrap, eval = FALSE}
inv_boot <- frame_invariance(
  fit,
  se_method = "bootstrap",
  boot_reps = 300,
  seed = 1
)
```

The conditional method returns raw and Holm-adjusted probabilities for the
location comparisons. Its standardised-infit and slope-ratio columns are
descriptive. The bootstrap returns discrimination probabilities as well and
applies Holm adjustment to the combined location and discrimination family.

In 300 null simulations with 500 persons per frame, the bootstrap SE ratios
were 1.00 for locations and 1.03 for log discrimination ratios; 95 per cent
coverage was 0.948 and 0.954, and combined Holm familywise error was 3.0 per
cent. A separate 2,000-replicate check found that probabilities based only on
the conditional standardised-infit comparison gave 7.1 per cent combined
familywise error. The conditional method therefore leaves those probabilities
undefined. At the same sample size, bootstrap power was 96.3 per cent for two
items shifted by one logit, but 9.6 per cent for two items made 1.5 times as
steep. Discrimination departures require appreciably more information.

The location contrasts are relative to the mean difference of the common
items. If a small number of items is shifted, centring also gives the
remaining items non-zero relative contrasts. The table identifies the pattern
of departures; item content or external anchors are needed to decide which
items provide the defensible reference.

## Refit after a departure

`resolve_frames()` gives an item a separate location in each frame. The item
continues to contribute to person measurement but no longer links those
frames. `drop_items()` removes it. Both functions refit the model and update
the frame, item and person estimates.
Each set must retain common items linking its groups' origins; a unit link
through another set is not sufficient.

```{r repair, eval = FALSE}
resolved <- resolve_frames(fit, "S1I02")
removed <- drop_items(fit, "S1I02")
```

The remaining common items and linked sets must still identify the frame
units. The functions preserve the original crossed factors, uncertainty
method and fitting controls, and return a result only when the revised model
converges.

## References

Follmann, D. (1988). Consistent estimation in the Rasch model based on
nonparametric margins. *Psychometrika*, 53, 553--562.
https://doi.org/10.1007/BF02294407

Humphry, S. M. (2005). *Maintaining a Common Arbitrary Unit in Social
Measurement*. PhD thesis, Murdoch University.

Humphry, S. M., and Andrich, D. (2008). Understanding the unit in the Rasch
model. *Journal of Applied Measurement*, 9(3), 249--264.

Rasch, G. (1961). On general laws and the meaning of measurement in
psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical
Statistics and Probability* (Vol. 4, pp. 321--333). University of California
Press.
